---
title: "Bioreactor_Stats_Report"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

check.packages <- function(package){
  new.package <- package[!(package %in% installed.packages()[, "Package"])]
  if (length(new.package)) 
    install.packages(new.package, dependencies = TRUE)
  sapply(package, require, character.only = TRUE)
}

packages<-c("tidyverse", "readxl", "car", "MASS", "lme4", "arm", "here", "fitdistrplus", "emmeans")

check.packages(packages)


```

# Data Input and Cleanup

Read in PFAS and methane production data from database file. Also stored analytical LOQs and reactor coding by experimental factor. 

```{r input}
raw.inputs <- read_excel('Reactor Raws/Johnsie Clean Data 2.xlsx', sheet = 2)

CH4.inputs <- read_excel('Reactor Raws/Johnsie Clean Data 2.xlsx', sheet = 1)

LoQs <- read_excel('Reactor Raws/Johnsie Clean Data 2.xlsx', sheet = 3)
LOQs <- structure(as.numeric(LoQs), names=colnames(LoQs))

codings <- read_excel('Reactor Raws/Johnsie Clean Data 2.xlsx', sheet = 5) %>%
  mutate(crossed_factors = paste0(Season,"_",biotype)) %>%
  group_by(crossed_factors)%>%
  mutate(dummy_rep = seq_along(crossed_factors))

codings

as.tibble(LOQs)
```

Single point imputation of missing data. Non-measurements days were dropped from the dataframe, values reported below LOQ were inputted at LoQ/sqrt(2), values as zero or ND were set at LOQ/10. These values will be replaced by later multipoint inputation on a per-compound basis if/when necessary. 

```{r cleaning}
clean_data <- raw.inputs %>%
  dplyr::filter(!is.na(PFBA)) %>%
  dplyr::select(-`Sample Date`, -`Date Analyzed`)

for (i in 1:length(LOQs)) {
  n = i+2
  clean_data[which(clean_data[,(n)] == "<LOQ"),(n)] <- LOQs[i]/sqrt(2) 
  clean_data[which(clean_data[,(n)] == "LOQ"),(n)] <- LOQs[i]/sqrt(2) 
  clean_data[which(clean_data[,(n)] == "ND"),(n)] <- LOQs[i]/10
  clean_data[which(clean_data[,(n)] == "0"),(n)] <- LOQs[i]/10
}

clean_data2 <- clean_data %>%
  mutate_at(colnames(clean_data[2:72]), as.numeric)
```

Methane production levels used to define reactor "phases"."Lag" phase is the initial phase with production < 0.2, "decline" phase is the final phase with production < 0.2. "Peak" phase is the period of microbe activity. Phases will be used if strict Days.Elapsed models fail. 

``` {r methane}
CH4.phase <- CH4.inputs %>%
  group_by(Reactor) %>%
  summarize(max.CH4 = max(CH4)) %>%
  left_join(CH4.inputs) %>%
  mutate(peak.day = ifelse((CH4 == max.CH4), Days.Elapsed,NA)) %>%
  dplyr::filter(!is.na(peak.day)) %>%
  dplyr::select(peak.day,Reactor)%>%
  full_join(CH4.inputs) %>%
  mutate(phase = ifelse(CH4 < 0.2 & Days.Elapsed < peak.day, "lag", "peak"),
         phase = ifelse(CH4 < 0.2 & Days.Elapsed > peak.day, "decline", phase)) %>%
  dplyr::select(Days.Elapsed,Reactor,phase) %>%
  dplyr::filter(!is.na(phase))
  
CH4.phase
```
  
Each abiotic reactor was assigned a biotic reactor from its "Season" for the purpose of dummy coding phase (acts as a time variable if we do phase based comparisons)
  
``` {r phases}
phase_factoring <- function(df,reactor.list,lag.knot,decline.knot) {
  df %>% mutate(phase = case_when(Reactor %in% reactor.list & Days.Elapsed <= lag.knot ~ "lag",
                                  Reactor %in% reactor.list & Days.Elapsed > lag.knot & Days.Elapsed < decline.knot ~ "peak",
                                  Reactor %in% reactor.list & Days.Elapsed >= decline.knot ~ "decline",
                                  TRUE ~ as.character(phase)))
                
}

## assigned manually because I am too tired to write a general function
final_raw <- clean_data2 %>% full_join(codings, by = "Reactor") %>% 
  mutate(phase = "waiting") %>%
  phase_factoring(c("P2"), 74, 220)%>%
  phase_factoring(c("P3"), 311, 458)%>%
  phase_factoring(c("P20"), 75, 335)%>%
  phase_factoring(c("P21"), 67, 257)%>%
  phase_factoring(c("P24"), 20, 140)%>%
  phase_factoring(c("P25"), 33, 208)%>%
  phase_factoring(c("P25"), 33, 208)%>%
  phase_factoring(c("P7"), 20, 132)%>%
  phase_factoring(c("P6"), 54, 193)%>%
  phase_factoring(c("P4","P5"), 74, 220) %>%
  phase_factoring(c("P22","P23"), 109, 296)%>%
  phase_factoring(c("P26","P27"),26, 174) %>%
  phase_factoring(c("P8","P9"), 37, 163) %>%
  mutate_at(c("Reactor", "Season", "biotype","phase"), as.factor)
  
final_raw
```

# Exploratory Data Analysis

## Overall Summaries

Calculate some summary stats for each PFAS measurement across all reactors to see what we are working with. 

``` {r summary_stats}
PFASs <- colnames(final_raw)[3:72]

means <- final_raw[PFASs] %>%
  summarize_all(.funs = funs(mean))
median <- final_raw[PFASs] %>%
  summarize_all(.funs = funs(median))
sd <- final_raw[PFASs] %>%
  summarize_all(.funs = funs(sd))
min <- final_raw[PFASs] %>%
  summarize_all(.funs = funs(min))
max <- final_raw[PFASs] %>%
  summarize_all(.funs = funs(max))

summaries <- bind_rows(means,median,sd,min,max) %>% as.data.frame()
rownames(summaries) <- c("mean","median","sd","min","max")

round(as.tibble(summaries), 2)
```

Some compounds show basically zero variation, others aren't detected hardly at all. Analysis will proceed on the "interesting" subset of compounds selected by Lang (DOI: 10.1021/acs.est.6b05005) with an additional filter to require a max detected concentration of 10. Other samples have too little variation to model. 

```{r PFAS_subset}
PFASs_subset <- c("PFBA","PFPeA","PFHxA","PFHpA","PFOA","PFNA","PFDA",
  "FHEA","FHUEA","FPePA","FOEA","FHpPA",
  "PFBS","PFHxS","PFOS",
  "MeFBSAA","FTS.6.2","diPAP.6.2","diPAP.6.2.8.2","diPAP.8.2")

subset2 <- PFASs_subset[which(summaries["max",PFASs_subset] > 10)]

working_raws <- final_raw[subset2] %>%
  cbind(final_raw[c("Days.Elapsed","dil","Reactor","Season","biotype","phase")])

summary_plot <- working_raws %>%
  gather(one_of(subset2), key = "PFAS", value = "raw_measure") %>%
  mutate(conc = raw_measure * dil)

as.tibble(summary_plot)
```

```{r summary_graphs}
##### LEGACY COMPOUNDS #####

legacy_summary <- filter(summary_plot, PFAS %in% c("PFBA","PFDA","PFOA","PFHpA","PFHxA","PFNA","PFOA","PFOS","PFPeA", "PFHxS"))

ggplot(filter(legacy_summary)) +
  theme_minimal()+
  geom_point(aes(x = Days.Elapsed, y = conc, group = Reactor), size = 0.5) +
  geom_line(aes(x = Days.Elapsed, y = conc, group = Reactor, color = biotype)) +
  scale_y_log10() +
  annotation_logticks(sides = "l")+
  facet_wrap(~PFAS) +
  labs(title ="Concentration v. Time for Legacy Compunds",
       x ="Days Elapsed Since Reactor Start",
       y ="Dilution Corrected PFAS Concentration (ng/L)")


##### F COMPOUNDS #####
telomer_summary <- filter(summary_plot, PFAS %in% c("FHEA","FHUEA","FPePA","FHpPA","FTS.6.2"))

ggplot(filter(telomer_summary)) +
  theme_minimal()+
  geom_point(aes(x = Days.Elapsed, y = conc, group = Reactor), size = 0.5) +
  geom_line(aes(x = Days.Elapsed, y = conc, group = Reactor, color = biotype)) +
  scale_y_log10() +
  annotation_logticks(sides = "l")+
  facet_wrap(~PFAS) +
  labs(title ="Concentration v. Time for Fluorotelomer Compunds",
       x ="Days Elapsed Since Reactor Start",
       y ="Dilution Corrected PFAS Concentration (ng/L)")



###### OTHER COMPOUNDS #####

other_summary <- filter(summary_plot, ! PFAS %in% telomer_summary$PFAS & ! PFAS %in% legacy_summary$PFAS)

ggplot(filter(other_summary)) +
  theme_minimal()+
  geom_point(aes(x = Days.Elapsed, y = conc, group = Reactor), size = 0.5) +
  geom_line(aes(x = Days.Elapsed, y = conc, group = Reactor, color = biotype)) +
  scale_y_log10() +
  annotation_logticks(sides = "l")+
  facet_wrap(~PFAS)+
  labs(title ="Concentration v. Time for detected PAPs and  Compunds",
       x ="Days Elapsed Since Reactor Start",
       y ="Dilution Corrected PFAS Concentration (ng/L)") 
```

## Legacy Compound Modeling

### PFBA

Select PFBA for individual compound analysis. Exploratory graphs.
``` {r PFBA_selecting}
chosen_PFAS <- "PFBA"

single_PFAS <- filter(summary_plot, PFAS == chosen_PFAS)

levels(single_PFAS$phase) <- c("lag", "peak","decline")

ggplot(single_PFAS) +
  geom_hline(yintercept = LOQs[chosen_PFAS], linetype = 2)+
  theme_minimal()+
  geom_point(aes(x = Days.Elapsed, y = raw_measure, group = Reactor), size = 0.5) +
  geom_line(aes(x = Days.Elapsed, y = raw_measure, group = Reactor, color = biotype)) +
  scale_y_log10() +
  annotation_logticks(sides = "l")+
  facet_wrap(~Reactor) +
  labs(title ="Concentration v. Time for Legacy Compunds",
       x ="Days Elapsed Since Reactor Start",
       y ="Concentration (ng/L) + LOQ(dashed)")

qqnorm(single_PFAS$raw_measure)
```

Single value inputation is pretty clear from the QQ plot. LOQ is very high (~60ng/L) so 80% of measures should be some inputed value. 

#### PFBA Multi-Value Inputation

Fit a log-normal distribution based on truncated distribution with censored data (using _fitdistrplus_)

``` {r PFBA_fitting}
fitting_values <- tibble(raw_measure = single_PFAS$raw_measure) %>%
  mutate(index1 = seq_len(nrow(.))) %>%
  sample_n(size = nrow(.)) %>%
  mutate(index2 = seq_len(nrow(.)))

lower_bound <- round(LOQs[chosen_PFAS]/10, 5)
sqrtloq <- round(LOQs[chosen_PFAS]/sqrt(2), 5)
cutoff <- round(LOQs[chosen_PFAS],5)

fitting_values <-  fitting_values %>%
  arrange(raw_measure) %>%
  mutate(raw_measure = round(raw_measure, 5)) %>%
  mutate(left = case_when(raw_measure <= lower_bound ~ as.numeric(NA),
                          raw_measure <= cutoff ~ lower_bound,
                          TRUE ~ raw_measure)) %>%
  mutate(right = case_when(raw_measure == lower_bound ~ lower_bound,
                           raw_measure <= cutoff ~ cutoff,
                           TRUE ~ raw_measure)) %>%
  arrange(raw_measure)

censdata <- fitting_values[c("left","right")] %>% as.data.frame() %>% arrange(left)

trunc_dist <- fitdistcens(log(censdata), distr = "norm")

plot(trunc_dist)
```

Reasonable fit I guess, we really are inputing 80% of the data so I am not sure it is actually meaningful.

#### PFBA Multi-Point Inputation
Going to do multi-point inputation instead of single point inputation. Assigned a random index to each censored data point, then fit with quantiles from the distribution fit above. 

``` {r PFBA_inputation, fig.cap = "Inputed data points overlaid on line plot of original reported values"}

trunc_quantiles <- unlist(quantile(trunc_dist, probs = seq(0,100,by=(100/(max(fitting_values$index1)+2)))/100)[1])[-1]

lowestcens <- sum(fitting_values$raw_measure <= lower_bound)
lowercens <- sum(fitting_values$raw_measure <= cutoff) - lowestcens

lowestcens_quatiles <- trunc_quantiles[1:lowestcens]
lowercens_quantiles <- trunc_quantiles[(lowestcens+1):(lowestcens+lowercens)]

inputions <- c(lowestcens_quatiles,lowercens_quantiles)

length(inputions) <- nrow(fitting_values)

inputed_frame <- cbind(fitting_values, inputions) %>%
  arrange(index1) %>%
  mutate(quantiles = ifelse(is.na(inputions), log(raw_measure), inputions)) %>%
  dplyr::select(quantiles) %>%
  cbind(single_PFAS) %>%
  mutate(conc = ifelse(raw_measure < LOQs[chosen_PFAS], exp(quantiles) * dil, raw_measure*dil)) %>%
  left_join(codings) %>%
  mutate(scaled_days = Days.Elapsed/100)

graph <- ggplot(inputed_frame, aes(x = Days.Elapsed, group = Reactor)) +
  theme_classic()+
  guides(color = FALSE, fill =FALSE, linetype = FALSE)+
  geom_hline(yintercept = LOQs[chosen_PFAS], linetype = 2) +
  geom_point(aes(y = conc,
                 color = as.factor(dummy_rep))) +
  geom_line(aes(y = raw_measure*dil,
                color = as.factor(dummy_rep))) +
  scale_color_manual(values = c("blue","green"))+
  facet_grid(biotype~Season) + 
  labs(title ="Concentration v. Time for Bioreactor Replicates",
       x ="Days Elapsed Since Reactor Start",
       y ="Concentration (ng/L) + LOQ(dashed)")

plot(graph)
```

Added noise should allow for modeling.

#### PFBA Modeling

Building linear mixed effect models to account for experimental factors and nesting of Reactor %in% Season. Using scaled_days (Days/100) to allow for model convergence, this effect can be reversed during model interpretation.

```{r PFBa_modeling}
model <- lmer(data = inputed_frame,
              log(conc) ~ scaled_days * biotype + Season + (scaled_days|Reactor))

ggplot(inputed_frame, aes(x = Days.Elapsed, group = Reactor)) +
  theme_classic()+
  guides(color = FALSE, fill =FALSE, linetype = FALSE)+
  geom_hline(yintercept = LOQs[chosen_PFAS], linetype = 2) +
  geom_point(aes(y = conc,
                 color = as.factor(dummy_rep))) +
  geom_line(aes(y = exp(predict(model)),
                color = as.factor(dummy_rep))) +
  facet_grid(biotype~Season) + 
  labs(title ="Measured vs. Modeled Reactor Concentrations",
       x ="Days Elapsed Since Reactor Start",
       y ="Concentration (ng/L) + LOQ(dashed)")

```

#### Post-Hoc Comparison Testing

Model summary:

```{r PFBA_posthoc_1}

summary(model)
```

Type II Wald p-values:

```{r PFBA_Wald}
afex::mixed(model, data = inputed_frame)

```



``` {r PFBA_emmeans}

PFBA.emtrends <- emtrends(model, ~ Season + biotype + scaled_days:biotype, var = "scaled_days", transform = "response")
emmeans.model <- emmeans(model, ~Season + biotype + scaled_days:biotype, transform = "response")

CLD(PFBA.emtrends)
    
CLD(emmeans.model)

plot(PFBA.emtrends, by = "Season", comparison = TRUE)

plot(emtrends(model, ~ Season + biotype, var = "scaled_days", transform = "response"), by = "biotype", comparison = TRUE)

plot(emmeans(model, ~ Season + biotype , transform = "response"), by = "biotype", comparison = TRUE)

#emmip(model, ~ Season | biotype, transform = "response") +
#  theme_bw()+
#  geom_hline(yintercept = LOQs[chosen_PFAS], linetype = 2)

```

### PFPeA

Select PFPeA for individual compound analysis. Exploratory graphs.
``` {r PFPeA_selecting}
chosen_PFAS <- "PFPeA"

single_PFAS <- filter(summary_plot, PFAS == chosen_PFAS)

levels(single_PFAS$phase) <- c("lag", "peak","decline")

ggplot(single_PFAS) +
  geom_hline(yintercept = LOQs[chosen_PFAS], linetype = 2)+
  theme_minimal()+
  geom_point(aes(x = Days.Elapsed, y = raw_measure, group = Reactor), size = 0.5) +
  geom_line(aes(x = Days.Elapsed, y = raw_measure, group = Reactor, color = biotype)) +
  scale_y_log10() +
  annotation_logticks(sides = "l")+
  facet_wrap(~Reactor) +
  labs(title ="Concentration v. Time for Legacy Compunds",
       x ="Days Elapsed Since Reactor Start",
       y ="Concentration (ng/L) + LOQ(dashed)")

qqnorm(single_PFAS$raw_measure)
```

Single value inputation is pretty clear from the QQ plot. LOQ is slightly high (30ng/L) so ~50% of measures should be some inputed value. 

#### PFPeA Multi-Value Inputation

Fit a log-normal distribution based on truncated distribution with censored data (using _fitdistrplus_)

``` {r PFPeA_fitting}
fitting_values <- tibble(raw_measure = single_PFAS$raw_measure) %>%
  mutate(index1 = seq_len(nrow(.))) %>%
  sample_n(size = nrow(.)) %>%
  mutate(index2 = seq_len(nrow(.)))

lower_bound <- round(LOQs[chosen_PFAS]/10, 5)
sqrtloq <- round(LOQs[chosen_PFAS]/sqrt(2), 5)
cutoff <- round(LOQs[chosen_PFAS],5)

fitting_values <-  fitting_values %>%
  arrange(raw_measure) %>%
  mutate(raw_measure = round(raw_measure, 5)) %>%
  mutate(left = case_when(raw_measure <= lower_bound ~ as.numeric(NA),
                          raw_measure <= cutoff ~ lower_bound,
                          TRUE ~ raw_measure)) %>%
  mutate(right = case_when(raw_measure == lower_bound ~ lower_bound,
                           raw_measure <= cutoff ~ cutoff,
                           TRUE ~ raw_measure)) %>%
  arrange(raw_measure)

censdata <- fitting_values[c("left","right")] %>% as.data.frame() %>% arrange(left)

trunc_dist <- fitdistcens(log(censdata), distr = "norm")

plot(trunc_dist)
```

Fairly long tail on the QQ-plot high side, we really are inputing 50% of the data so I am not sure it is actually meaningful.

#### PFPeA Multi-Point Inputation
Going to do multi-point inputation instead of single point inputation. Assigned a random index to each censored data point, then fit with quantiles from the distribution fit above. 

``` {r PFPeA_inputation, fig.cap = "Inputed data points overlaid on line plot of original reported values"}

trunc_quantiles <- unlist(quantile(trunc_dist, probs = seq(0,100,by=(100/(max(fitting_values$index1)+2)))/100)[1])[-1]

lowestcens <- sum(fitting_values$raw_measure <= lower_bound)
lowercens <- sum(fitting_values$raw_measure <= cutoff) - lowestcens

lowestcens_quatiles <- trunc_quantiles[1:lowestcens]
lowercens_quantiles <- trunc_quantiles[(lowestcens+1):(lowestcens+lowercens)]

inputions <- c(lowestcens_quatiles,lowercens_quantiles)

length(inputions) <- nrow(fitting_values)

inputed_frame <- cbind(fitting_values, inputions) %>%
  arrange(index1) %>%
  mutate(quantiles = ifelse(is.na(inputions), log(raw_measure), inputions)) %>%
  dplyr::select(quantiles) %>%
  cbind(single_PFAS) %>%
  mutate(conc = ifelse(raw_measure < LOQs[chosen_PFAS], exp(quantiles) * dil, raw_measure*dil)) %>%
  left_join(codings) %>%
  mutate(scaled_days = Days.Elapsed/100)

graph <- ggplot(inputed_frame, aes(x = Days.Elapsed, group = Reactor)) +
  theme_classic()+
  guides(color = FALSE, fill =FALSE, linetype = FALSE)+
  geom_hline(yintercept = LOQs[chosen_PFAS], linetype = 2) +
  geom_point(aes(y = conc,
                 color = as.factor(dummy_rep))) +
  geom_line(aes(y = raw_measure*dil,
                color = as.factor(dummy_rep))) +
  scale_color_manual(values = c("blue","green"))+
  facet_grid(biotype~Season) + 
  labs(title ="Concentration v. Time for Bioreactor Replicates",
       x ="Days Elapsed Since Reactor Start",
       y ="Concentration (ng/L) + LOQ(dashed)")

plot(graph)
```

Added noise should allow for modeling.

#### PFPeA Modeling

Building linear mixed effect models to account for experimental factors and nesting of Reactor %in% Season. Using scaled_days (Days/100) to allow for model convergence, this effect can be reversed during model interpretation.

```{r PFPeA_modeling}
model <- lmer(data = inputed_frame,
              log(conc) ~ scaled_days * biotype + Season + (1|Reactor))

ggplot(inputed_frame, aes(x = Days.Elapsed, group = Reactor)) +
  theme_classic()+
  guides(color = FALSE, fill =FALSE, linetype = FALSE)+
  geom_hline(yintercept = LOQs[chosen_PFAS], linetype = 2) +
  geom_point(aes(y = conc,
                 color = as.factor(dummy_rep))) +
  geom_line(aes(y = exp(predict(model)),
                color = as.factor(dummy_rep))) +
  facet_grid(biotype~Season) + 
  labs(title ="Measured vs. Modeled Reactor Concentrations",
       x ="Days Elapsed Since Reactor Start",
       y ="Concentration (ng/L) + LOQ(dashed)")

```

#### Post-Hoc Comparison Testing

Model summary:

```{r PFPeA_posthoc_1}

summary(model)
```

Type II Wald p-values:

```{r PFPeA_Wald}
afex::mixed(model, data = inputed_frame)

```



``` {r PFPeA_emmeans}

PFPeA.emtrends.model <- emtrends(model, ~ Season + biotype + scaled_days:biotype, var = "scaled_days", transform = "response")
PFPeA.emmeans.model <- emmeans(model, ~Season + biotype + scaled_days:biotype, transform = "response")

CLD(PFPeA.emtrends.model)
    
CLD(PFPeA.emmeans.model)

plot(PFPeA.emtrends.model, by = "Season", comparison = TRUE)

plot(PFPeA.emtrends.model, by = "biotype", comparison = TRUE)

plot(PFPeA.emmeans.model, by = "biotype", comparison = TRUE)

#emmip(model, ~ Season | biotype, transform = "response") +
#  theme_bw()+
#  geom_hline(yintercept = LOQs[chosen_PFAS], linetype = 2)

```
### PFHxA

Select PFHxA for individual compound analysis. Exploratory graphs.
``` {r PFHxA_selecting}
chosen_PFAS <- "PFHxA"

single_PFAS <- filter(summary_plot, PFAS == chosen_PFAS)

levels(single_PFAS$phase) <- c("lag", "peak","decline")

ggplot(single_PFAS) +
  geom_hline(yintercept = LOQs[chosen_PFAS], linetype = 2)+
  theme_minimal()+
  geom_point(aes(x = Days.Elapsed, y = raw_measure, group = Reactor), size = 0.5) +
  geom_line(aes(x = Days.Elapsed, y = raw_measure, group = Reactor, color = biotype)) +
  scale_y_log10() +
  annotation_logticks(sides = "l")+
  facet_wrap(~Reactor) +
  labs(title ="Concentration v. Time for Legacy Compunds",
       x ="Days Elapsed Since Reactor Start",
       y ="Concentration (ng/L) + LOQ(dashed)")

unscaled_qq <- function() {
qqnorm(single_PFAS$raw_measure)
qqline(single_PFAS$raw_measure)
}

unscaled_qq()

log_qq <- function() {
  qqnorm(log(single_PFAS$raw_measure))
qqline(log(single_PFAS$raw_measure))
}

log_qq()
```

Single value inputation is pretty clear from the QQ plot. log-scaled qq looks fairly normal, we can head straight into analysis. 

#### PFHxA Multi-Value Inputation

Fit a log-normal distribution based on truncated distribution with censored data (using _fitdistrplus_)

``` {r PFHxA_fitting}
fitting_values <- tibble(raw_measure = single_PFAS$raw_measure) %>%
  mutate(index1 = seq_len(nrow(.))) %>%
  sample_n(size = nrow(.)) %>%
  mutate(index2 = seq_len(nrow(.)))

lower_bound <- round(LOQs[chosen_PFAS]/10, 5)
sqrtloq <- round(LOQs[chosen_PFAS]/sqrt(2), 5)
cutoff <- round(LOQs[chosen_PFAS],5)

fitting_values <-  fitting_values %>%
  arrange(raw_measure) %>%
  mutate(raw_measure = round(raw_measure, 5)) %>%
  mutate(left = case_when(raw_measure <= lower_bound ~ as.numeric(NA),
                          raw_measure <= cutoff ~ lower_bound,
                          TRUE ~ raw_measure)) %>%
  mutate(right = case_when(raw_measure == lower_bound ~ lower_bound,
                           raw_measure <= cutoff ~ cutoff,
                           TRUE ~ raw_measure)) %>%
  arrange(raw_measure)

censdata <- fitting_values[c("left","right")] %>% as.data.frame() %>% arrange(left)

trunc_dist <- fitdistcens(log(censdata), distr = "norm")

plot(trunc_dist)
```

No censored data so fits a log-normal dist pretty cleanly as we saw from the previous QQ-Plots

``` {r PFHxA_inputation, fig.cap = "Inputed data points overlaid on line plot of original reported values", echo =FALSE}

trunc_quantiles <- unlist(quantile(trunc_dist, probs = seq(0,100,by=(100/(max(fitting_values$index1)+2)))/100)[1])[-1]

lowestcens <- sum(fitting_values$raw_measure <= lower_bound)
lowercens <- sum(fitting_values$raw_measure <= cutoff) - lowestcens

lowestcens_quatiles <- trunc_quantiles[1:lowestcens]
lowercens_quantiles <- trunc_quantiles[(lowestcens+1):(lowestcens+lowercens)]

inputions <- c(lowestcens_quatiles,lowercens_quantiles)

length(inputions) <- nrow(fitting_values)

inputed_frame <- cbind(fitting_values, inputions) %>%
  arrange(index1) %>%
  mutate(quantiles = ifelse(is.na(inputions), log(raw_measure), inputions)) %>%
  dplyr::select(quantiles) %>%
  cbind(single_PFAS) %>%
  mutate(conc = ifelse(raw_measure < LOQs[chosen_PFAS], exp(quantiles) * dil, raw_measure*dil)) %>%
  left_join(codings) %>%
  mutate(scaled_days = Days.Elapsed/100)

graph <- ggplot(inputed_frame, aes(x = Days.Elapsed, group = Reactor)) +
  theme_classic()+
  guides(color = FALSE, fill =FALSE, linetype = FALSE)+
  geom_hline(yintercept = LOQs[chosen_PFAS], linetype = 2) +
  geom_point(aes(y = conc,
                 color = as.factor(dummy_rep))) +
  geom_line(aes(y = raw_measure*dil,
                color = as.factor(dummy_rep))) +
  scale_color_manual(values = c("blue","green"))+
  facet_grid(biotype~Season) + 
  labs(title ="Concentration v. Time for Bioreactor Replicates",
       x ="Days Elapsed Since Reactor Start",
       y ="Concentration (ng/L) + LOQ(dashed)")

plot(graph)
```



#### PFHxA Modeling

Building linear mixed effect models to account for experimental factors and nesting of Reactor %in% Season. Using scaled_days (Days/100) to allow for model convergence, this effect can be reversed during model interpretation.

```{r PFHxA_modeling}
model <- lmer(data = inputed_frame,
              log(conc) ~ scaled_days * biotype + Season + (1|Reactor))

ggplot(inputed_frame, aes(x = Days.Elapsed, group = Reactor)) +
  theme_classic()+
  guides(color = FALSE, fill =FALSE, linetype = FALSE)+
  geom_hline(yintercept = LOQs[chosen_PFAS], linetype = 2) +
  geom_point(aes(y = conc,
                 color = as.factor(dummy_rep))) +
  geom_line(aes(y = exp(predict(model)),
                color = as.factor(dummy_rep))) +
  facet_grid(biotype~Season) + 
  labs(title ="Measured vs. Modeled Reactor Concentrations",
       x ="Days Elapsed Since Reactor Start",
       y ="Concentration (ng/L) + LOQ(dashed)")

```

#### Post-Hoc Comparison Testing

Model summary:

```{r PFHxA_posthoc_1}

summary(model)
```

Type II Wald p-values:

```{r PFHxA_Wald}
afex::mixed(model, data = inputed_frame)

```

No statistically significant differences based on Reactor, Season, or time. Only significant parameter is likely physical release _(?)_ over time

### PFHpA

Select PFHpA for individual compound analysis. Exploratory graphs.
``` {r PFHpA_selecting}
chosen_PFAS <- "PFHpA"

single_PFAS <- filter(summary_plot, PFAS == chosen_PFAS)

levels(single_PFAS$phase) <- c("lag", "peak","decline")

ggplot(single_PFAS) +
  geom_hline(yintercept = LOQs[chosen_PFAS], linetype = 2)+
  theme_minimal()+
  geom_point(aes(x = Days.Elapsed, y = raw_measure, group = Reactor), size = 0.5) +
  geom_line(aes(x = Days.Elapsed, y = raw_measure, group = Reactor, color = biotype)) +
  scale_y_log10() +
  annotation_logticks(sides = "l")+
  facet_wrap(~Reactor) +
  labs(title ="Concentration v. Time for Legacy Compunds",
       x ="Days Elapsed Since Reactor Start",
       y ="Concentration (ng/L) + LOQ(dashed)")

unscaled_qq <- function() {
qqnorm(single_PFAS$raw_measure)
qqline(single_PFAS$raw_measure)
}

unscaled_qq()

log_qq <- function() {
  qqnorm(log(single_PFAS$raw_measure))
qqline(log(single_PFAS$raw_measure))
}

log_qq()
```

QQ plot almost normal, log-scaling clears upper end long tail. Using log-scaled data for consistent modeling.  


``` {r PFHpA_inputation, echo = FALSE}

trunc_quantiles <- unlist(quantile(trunc_dist, probs = seq(0,100,by=(100/(max(fitting_values$index1)+2)))/100)[1])[-1]

lowestcens <- sum(fitting_values$raw_measure <= lower_bound)
lowercens <- sum(fitting_values$raw_measure <= cutoff) - lowestcens

lowestcens_quatiles <- trunc_quantiles[1:lowestcens]
lowercens_quantiles <- trunc_quantiles[(lowestcens+1):(lowestcens+lowercens)]

inputions <- c(lowestcens_quatiles,lowercens_quantiles)

length(inputions) <- nrow(fitting_values)

inputed_frame <- cbind(fitting_values, inputions) %>%
  arrange(index1) %>%
  mutate(quantiles = ifelse(is.na(inputions), log(raw_measure), inputions)) %>%
  dplyr::select(quantiles) %>%
  cbind(single_PFAS) %>%
  mutate(conc = ifelse(raw_measure < LOQs[chosen_PFAS], exp(quantiles) * dil, raw_measure*dil)) %>%
  left_join(codings) %>%
  mutate(scaled_days = Days.Elapsed/100)

graph <- ggplot(inputed_frame, aes(x = Days.Elapsed, group = Reactor)) +
  theme_classic()+
  guides(color = FALSE, fill =FALSE, linetype = FALSE)+
  geom_hline(yintercept = LOQs[chosen_PFAS], linetype = 2) +
  geom_point(aes(y = conc,
                 color = as.factor(dummy_rep))) +
  geom_line(aes(y = raw_measure*dil,
                color = as.factor(dummy_rep))) +
  scale_color_manual(values = c("blue","green"))+
  facet_grid(biotype~Season) + 
  labs(title ="Concentration v. Time for Bioreactor Replicates",
       x ="Days Elapsed Since Reactor Start",
       y ="Concentration (ng/L) + LOQ(dashed)")

plot(graph)
```

#### PFHpA Modeling

Building linear mixed effect models to account for experimental factors and nesting of Reactor %in% Season. Using scaled_days (Days/100) to allow for model convergence, this effect can be reversed during model interpretation.

```{r PFHpA_modeling}
model <- lmer(data = inputed_frame,
              log(conc) ~ scaled_days * biotype + (1|Season/Reactor))

ggplot(inputed_frame, aes(x = Days.Elapsed, group = Reactor)) +
  theme_classic()+
  guides(color = FALSE, fill =FALSE, linetype = FALSE)+
  geom_hline(yintercept = LOQs[chosen_PFAS], linetype = 2) +
  geom_point(aes(y = conc,
                 color = as.factor(dummy_rep))) +
  geom_line(aes(y = exp(predict(model)),
                color = as.factor(dummy_rep))) +
  facet_grid(biotype~Season) + 
  labs(title ="Measured vs. Modeled Reactor Concentrations",
       x ="Days Elapsed Since Reactor Start",
       y ="Concentration (ng/L) + LOQ(dashed)")

```

#### Post-Hoc Comparison Testing

Model summary:

```{r PFHpA_posthoc_1}

summary(model)
```

Type II Wald p-values:

```{r PFHpA_Wald}
afex::mixed(model, data = inputed_frame)

```



``` {r PFHpA_emmeans}

PFHpA.emtrends.model <- emtrends(model, ~ scaled_days + Season + biotype + scaled_days:biotype, var = "scaled_days", transform = "response")
PFHpA.emmeans.model <- emmeans(model, ~scaled_days + Season + biotype + scaled_days:biotype, transform = "response")

CLD(PFHpA.emtrends.model)
    
CLD(PFHpA.emmeans.model)

plot(PFHpA.emtrends.model, by = "Season", comparison = TRUE)

plot(PFHpA.emtrends.model, by = "biotype", comparison = TRUE)

plot(PFHpA.emmeans.model, by = "biotype", comparison = TRUE)

#emmip(model, ~ Season | biotype, transform = "response") +
#  theme_bw()+
#  geom_hline(yintercept = LOQs[chosen_PFAS], linetype = 2)

```